---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

# AdaStableNet

AdaStableNet estimates a constant dynamic matrix in the autonomous linear ODE

\[
\frac{dX(t)}{dt} = A X(t)
\]

from a noisy multivariate trajectory. It combines functional-data smoothing,
gradient-matching initialization, and profiled nonlinear least squares in a
real modal basis. By default, the profiled ODE branches are fitted to the
B-spline-reconstructed trajectory rather than directly to the noisy values.
The package returns unconstrained, Wald-screened, and
stability-constrained fits together with diagnostics and forecasting methods.

The data are conventional time-domain trajectories: each row of the main
interface is an ordered observation time and each column is a system state.
AdaStableNet does not treat observations as biological cells and does not infer
or use pseudotime.

## Installation

```r
# install.packages("devtools")
devtools::install_github(
  "LopezRicardo1/AdaStableNet",
  build_vignettes = TRUE
)
```

## Quick start

```{r setup, message=FALSE}
library(AdaStableNet)
```

The simulator defaults to the sparse construction used in the original
eigen-bound study:

```{r sparse-design}
study_system <- simulate_adastablenet(
  p = 16,
  n_time = 31,
  spectrum = "marginal",
  sigma = 0.03,
  seed = 777
)
c(Q_sparsity = study_system$Q_sparsity,
  A_sparsity = study_system$A_sparsity)
```

```{r example}
sim <- simulate_adastablenet(
  p = 3,
  n_time = 31,
  spectrum = "stable",
  sigma = 0.03,
  seed = 2026
)

fit <- FitAdaStableNet(
  sim$Y,
  sim$time,
  nbasis = 12,
  twoSE = FALSE,
  fit_ode2fd = TRUE,
  eigen_real_wald = FALSE,
  num_iter = 100,
  verbose = FALSE
)

fit
summary(fit)
```

The fitted dynamic matrix and future trajectory use standard R methods:

```{r methods}
A_hat <- coef(fit, branch = "stable")
future <- predict(fit, new_time = seq(0, 2, length.out = 101))
head(future)
```

`predict()` propagates the fitted ODE with a matrix exponential. The profiled
branches start from their fitted modal state; `branch = "two_stage"` starts
from the B-spline-reconstructed state at the first training time. Thus the
forecast is a smooth ODE solution, not a spline extrapolation and not a curve
anchored to the first noisy measurement.

## Model branches

- `unbounded` estimates modal real parts without a sign constraint.
- `wald` applies a modal Wald screen and sets real parts that are not
  distinguishable from zero to zero.
- `stable` constrains retained real parts to lie below `stability_margin`, which
  is zero by default. This is the stability-constrained optimization branch.

The three ideas should not be conflated. `AdaStableNet_WaldTest()` performs the
Wu et al. (2019) trajectory-sensitivity Wald test for individual entries of
the coefficient matrix. The `wald` branch instead screens modal real parts,
and the `stable` branch imposes the spectral constraint.

## Wald-sparsified interaction network

Use `AdaStableNet_WaldNetwork()` to convert coefficient tests into a directed
network. In `X'(t) = A X(t)`, entry `A[i, j]` is the edge from state `j` to
state `i`. BH adjustment is applied only to off-diagonal edges. Diagonal
self-dynamics are retained in `A_sparse` by default but are zero in the
adjacency and weighted network.

```{r wald-network}
network <- AdaStableNet_WaldNetwork(
  fit, sim$Y, sim$time,
  branch = "stable", method = "BH", alpha = 0.05,
  nsteps = 4, stability_horizon = 1, stability_n_grid = 11
)

network$adjacency       # binary directed graph, diagonal is zero
network$A_network       # selected signed off-diagonal weights
network$A_sparse        # ODE matrix: selected edges plus self-dynamics
head(network$edge_table)
network$stability       # stability is recomputed after thresholding
```

Elementwise thresholding can change the eigenvalues, so a stable fitted branch
does not automatically make `A_sparse` stable. Use the full fitted matrix for
the primary trajectory estimate and treat sparse-matrix forecasting as a
sensitivity analysis unless the post-sparsification check remains acceptable.

With the default margin, `stable` guarantees a nonpositive realized spectral
abscissa when the reconstructed modal loading is full rank. A nonnormal matrix
can still have transient growth. Inspect all three stability measures:

```r
stability_diagnostics(fit, branch = "stable", horizon = 2)
```

Single-trajectory identifiability is a property of `(A, x0)`. Install
`ode.ident` from `qiuxing/ode.ident` and inspect ICIS with:

```r
identifiability_diagnostics(sim$A, sim$x0)
```

## Optional Torch backend

The default `backend = "base"` has no Torch dependency and uses `optim()`.
For automatic differentiation, install R Torch and its LibTorch runtime once:

```r
install.packages("torch")
torch::install_torch()
```

Then select the backend when fitting:

```r
fit_torch <- FitAdaStableNet(
  sim$Y,
  sim$time,
  nbasis = 12,
  backend = "torch",
  torch_device = "auto",
  lr = 0.01,
  torch_refine = TRUE,
  num_iter = 100,
  verbose = FALSE
)

fit_torch$control$backend
fit_torch$AdaEigenStableNet$Eigen_Bound$diagnostics[
  c("device", "dtype", "gradient")
]
```

The Torch path uses float64 tensors, Adam optimization, and optional L-BFGS
refinement. `torch_device = "auto"` uses CUDA when it is available and otherwise
uses the CPU. Set `backend = "auto"` to use Torch only when both the R package
and its runtime are available; the resolved backend is recorded in the fit.

The package checks the rank and condition number of the fitted modal loading
matrix and reports the actual spectral abscissa of every reconstructed system
matrix. The simulator defaults to the sparse block-embedding construction from
the original eigen-bound study. Use `matrix_structure = "dense"` when an exact
loading-matrix condition number is the intended design. See
`vignette("simulation-study", package = "AdaStableNet")` for the reproducible
simulation design and compact benchmark.

## Package documentation and paper separation

The installed vignette contains a compact, reproducible simulation and network-
recovery example built from exported package functions:

```r
vignette("simulation-study", package = "AdaStableNet")
```

Large Monte Carlo drivers, real-data preparation, checkpoints, manuscript
figures, tables, and journal source files are intentionally maintained in a
separate paper-analysis repository. They are not installed with AdaStableNet.

## Scope

The current model assumes a constant, autonomous, homogeneous linear ODE with
all states observed and a complete diagonalizable real modal representation.
The modal and coefficient Wald calculations condition on the fitted initial
state. The coefficient calculation follows the trajectory-sensitivity Fisher
information of Wu et al. (2019); after stability-constrained boundary fitting,
its normal reference distribution should be checked by simulation.
