---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

# AdaStableNet

AdaStableNet estimates a constant dynamic matrix in the autonomous linear ODE

\[
\frac{dX(t)}{dt} = A X(t)
\]

from a noisy multivariate trajectory. It combines functional-data smoothing,
gradient-matching initialization, and profiled nonlinear least squares in a
real modal basis. The package returns unconstrained, Wald-screened, and
stability-constrained fits together with diagnostics and forecasting methods.

The data are conventional time-domain trajectories: each row of the main
interface is an ordered observation time and each column is a system state.
AdaStableNet does not treat observations as biological cells and does not infer
or use pseudotime.

## Installation

```r
# install.packages("devtools")
devtools::install_github(
  "LopezRicardo1/AdaStableNet",
  build_vignettes = TRUE
)
```

## Quick start

```{r setup, message=FALSE}
library(AdaStableNet)
```

The simulator defaults to the sparse construction used in the original
eigen-bound study:

```{r sparse-design}
study_system <- simulate_adastablenet(
  p = 16,
  n_time = 31,
  spectrum = "marginal",
  sigma = 0.03,
  seed = 777
)
c(Q_sparsity = study_system$Q_sparsity,
  A_sparsity = study_system$A_sparsity)
```

```{r example}
sim <- simulate_adastablenet(
  p = 3,
  n_time = 31,
  spectrum = "stable",
  sigma = 0.03,
  seed = 2026
)

fit <- FitAdaStableNet(
  sim$Y,
  sim$time,
  nbasis = 12,
  twoSE = FALSE,
  eigen_real_wald = FALSE,
  num_iter = 100,
  verbose = FALSE
)

fit
summary(fit)
```

The fitted dynamic matrix and future trajectory use standard R methods:

```{r methods}
A_hat <- coef(fit, branch = "stable")
future <- predict(fit, new_time = seq(0, 2, length.out = 101))
head(future)
```

## Model branches

- `unbounded` estimates modal real parts without a sign constraint.
- `wald` applies a modal Wald screen and sets real parts that are not
  distinguishable from zero to zero.
- `stable` constrains retained real parts to lie below `stability_margin`, which
  is zero by default. This is the stability-constrained optimization branch.

The three ideas should not be conflated. `AdaStableNet_WaldTest()` performs the
Wu et al. (2019) trajectory-sensitivity Wald test for individual entries of
the coefficient matrix. The `wald` branch instead screens modal real parts,
and the `stable` branch imposes the spectral constraint.

With the default margin, `stable` guarantees a nonpositive realized spectral
abscissa when the reconstructed modal loading is full rank. A nonnormal matrix
can still have transient growth. Inspect all three stability measures:

```r
stability_diagnostics(fit, branch = "stable", horizon = 2)
```

Single-trajectory identifiability is a property of `(A, x0)`. Install
`ode.ident` from `qiuxing/ode.ident` and inspect ICIS with:

```r
identifiability_diagnostics(sim$A, sim$x0)
```

## Optional Torch backend

The default `backend = "base"` has no Torch dependency and uses `optim()`.
For automatic differentiation, install R Torch and its LibTorch runtime once:

```r
install.packages("torch")
torch::install_torch()
```

Then select the backend when fitting:

```r
fit_torch <- FitAdaStableNet(
  sim$Y,
  sim$time,
  nbasis = 12,
  backend = "torch",
  torch_device = "auto",
  lr = 0.01,
  torch_refine = TRUE,
  num_iter = 100,
  verbose = FALSE
)

fit_torch$control$backend
fit_torch$AdaEigenStableNet$Eigen_Bound$diagnostics[
  c("device", "dtype", "gradient")
]
```

The Torch path uses float64 tensors, Adam optimization, and optional L-BFGS
refinement. `torch_device = "auto"` uses CUDA when it is available and otherwise
uses the CPU. Set `backend = "auto"` to use Torch only when both the R package
and its runtime are available; the resolved backend is recorded in the fit.

The package checks the rank and condition number of the fitted modal loading
matrix and reports the actual spectral abscissa of every reconstructed system
matrix. The simulator defaults to the sparse block-embedding construction from
the original eigen-bound study. Use `matrix_structure = "dense"` when an exact
loading-matrix condition number is the intended design. See
`vignette("simulation-study", package = "AdaStableNet")` for the reproducible
simulation design and compact benchmark.

## Source the simulation study

The installed runner uses the original sparse (p=16), seed-777 marginal design
and (p=15), seed-888 mixed design. This quick run uses two replications per
design:

```r
Sys.setenv(
  ADASTABLENET_N_REP = 2,
  ADASTABLENET_NOISE_SD = "0.30",
  ADASTABLENET_BACKEND = "base",
  ADASTABLENET_OUTPUT_DIR = file.path(
    getwd(), "adastablenet-simulation-results"
  )
)

source(system.file(
  "scripts", "run-simulation-study.R",
  package = "AdaStableNet"
))

simulation_system_diagnostics
head(simulation_results)
simulation_summary
```

For the paper run, set `ADASTABLENET_N_REP = 500`,
`ADASTABLENET_NOISE_SD = "0.05,0.15,0.30"`, and choose
`ADASTABLENET_BACKEND = "base"`, `"torch"`, or `"auto"` before sourcing the
same file. The runner checkpoints every completed replication, so sourcing it
again with the same output directory resumes the study.

## Scope

The current model assumes a constant, autonomous, homogeneous linear ODE with
all states observed and a complete diagonalizable real modal representation.
The modal and coefficient Wald calculations condition on the fitted initial
state. The coefficient calculation follows the trajectory-sensitivity Fisher
information of Wu et al. (2019); after stability-constrained boundary fitting,
its normal reference distribution should be checked by simulation.
