---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

# AdaStableNet

AdaStableNet estimates a constant dynamic matrix in the autonomous linear ODE

\[
\frac{dX(t)}{dt} = A X(t)
\]

from a noisy multivariate trajectory. It combines functional-data smoothing,
gradient-matching initialization, and profiled nonlinear least squares in a
real modal basis. The package returns unconstrained, Wald-screened, and
stability-constrained fits together with diagnostics and forecasting methods.

## Installation

```r
# install.packages("devtools")
devtools::install_github(
  "LopezRicardo1/AdaStableNet",
  build_vignettes = TRUE
)
```

## Quick start

```{r setup, message=FALSE}
library(AdaStableNet)
```

The simulator defaults to the sparse construction used in the original
eigen-bound study:

```{r sparse-design}
study_system <- simulate_adastablenet(
  p = 16,
  n_time = 31,
  spectrum = "marginal",
  sigma = 0.03,
  seed = 777
)
c(Q_sparsity = study_system$Q_sparsity,
  A_sparsity = study_system$A_sparsity)
```

```{r example}
sim <- simulate_adastablenet(
  p = 3,
  n_time = 31,
  spectrum = "stable",
  sigma = 0.03,
  seed = 2026
)

fit <- FitAdaStableNet(
  sim$Y,
  sim$time,
  nbasis = 12,
  twoSE = FALSE,
  eigen_real_wald = FALSE,
  num_iter = 100,
  verbose = FALSE
)

fit
summary(fit)
```

The fitted dynamic matrix and future trajectory use standard R methods:

```{r methods}
A_hat <- coef(fit, branch = "stable")
future <- predict(fit, new_time = seq(0, 2, length.out = 101))
head(future)
```

## Model branches

- `unbounded` estimates modal real parts without a sign constraint.
- `wald` sets modal real parts that are not distinguishable from zero to zero.
- `stable` constrains retained real parts to lie below `stability_margin`, which
  is zero by default.

The package checks the rank and condition number of the fitted modal loading
matrix and reports the actual spectral abscissa of every reconstructed system
matrix. The simulator defaults to the sparse block-embedding construction from
the original eigen-bound study. Use `matrix_structure = "dense"` when an exact
loading-matrix condition number is the intended design. See
`vignette("simulation-study", package = "AdaStableNet")` for the reproducible
simulation design and compact benchmark.

## Source the simulation study

The installed runner uses the original sparse (p=16), seed-777 marginal design
and (p=15), seed-888 mixed design. This quick run uses two replications per
design:

```r
Sys.setenv(
  ADASTABLENET_N_REP = 2,
  ADASTABLENET_NOISE_SD = "0.30",
  ADASTABLENET_OUTPUT_DIR = file.path(
    getwd(), "adastablenet-simulation-results"
  )
)

source(system.file(
  "scripts", "run-simulation-study.R",
  package = "AdaStableNet"
))

simulation_system_diagnostics
head(simulation_results)
simulation_summary
```

For the paper run, set `ADASTABLENET_N_REP = 500` and
`ADASTABLENET_NOISE_SD = "0.05,0.15,0.30"` before sourcing the same file. The
runner checkpoints every completed replication, so sourcing it again with the
same output directory resumes the study.

## Scope

The current model assumes a constant, autonomous, homogeneous linear ODE with
all states observed and a complete diagonalizable real modal representation.
The approximate Wald calculations condition on the fitted initial state.
