---
title: "Simulation Study for AdaStableNet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study for AdaStableNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
set.seed(2026)
```

## Purpose

This vignette provides an executable version of the simulation framework used
to assess AdaStableNet. The main estimand is the matrix (A) in

\[
X'(t)=A X(t), \qquad X(t)=\exp\{A(t-t_0)\}X(t_0).
\]

This is a standard time-domain ODE setting. Observations are multivariate state
measurements collected at ordered times; they are not cells, and the time
variable is observed directly rather than inferred as pseudotime.

The experiment separates three questions:

1. How accurately is (A) recovered?
2. Does the estimated system have the intended stability behavior?
3. How well does the estimate forecast beyond the observation interval?

Unlike an experiment that fits one trajectory and evaluates many initial
conditions, each Monte Carlo replication below generates new data and refits
every estimator. This measures estimator bias and variability rather than only
transfer across initial conditions.

```{r libraries, message=FALSE}
library(AdaStableNet)
```

## Data-generating systems

`simulate_adastablenet()` constructs a real modal matrix (J), a loading matrix
(Q), and (A=QJQ^{-1}). By default, `matrix_structure = "sparse"` reproduces
the eigen-bound notebook construction: two small orthogonal blocks are embedded
in an identity matrix, multiplied by a randomly permuted local rotation, and
entries below the median nonzero magnitude are set to zero. This produces a
sparse, non-orthogonal (Q) and a sparse (A). Set
`matrix_structure = "dense"` to use a dense loading matrix with an exact target
condition number.

The realized sparsity is recorded in `Q_sparsity` and `A_sparsity`. In very
small systems, the embedded blocks can cover most coordinates, so the realized
proportion of zeros in (A) may be lower than in the original (p=15) and (p=16)
designs.

The available spectral regimes are:

- `stable`: every real part is negative;
- `marginal`: all real parts are zero;
- `mixed`: marginal oscillations plus a stable real mode for odd dimensions;
  even dimensions alternate marginal and stable oscillatory modes;
- `near_boundary`: negative real parts close to zero;
- `unstable`: positive real parts, used as a misspecification check.

The two experiments in the original ODE eigen-bound notebook correspond most
closely to a marginal even-dimensional system and a mixed odd-dimensional
system. They remain important boundary cases because a strict stability
constraint can bias a truly marginal process unless the Wald screen retains a
zero real part.

```{r one-system}
sparse_system <- simulate_adastablenet(
  p = 16,
  n_time = 61,
  time_range = c(0, 2),
  spectrum = "marginal",
  sigma = 0.08,
  condition_number = 5,
  seed = 777
)

sort(eigen(sparse_system$A)$values)
sparse_system[c("matrix_structure", "Q_sparsity", "A_sparsity",
                "condition_number")]

example_system <- simulate_adastablenet(
  p = 4,
  n_time = 61,
  time_range = c(0, 2),
  spectrum = "mixed",
  sigma = 0.08,
  condition_number = 8,
  seed = 11
)
```

```{r sparsity-pattern, fig.cap="Nonzero pattern of the sparse system matrix A."}
image(
  x = seq_len(nrow(sparse_system$A)),
  y = seq_len(ncol(sparse_system$A)),
  z = abs(sparse_system$A) >= 1e-12,
  col = c("white", "#0072B2"),
  xlab = "Row", ylab = "Column", axes = FALSE
)
axis(1)
axis(2)
box()
```

Only observations on ([0,1]) are used for estimation; ((1,2]) is reserved for
future prediction.

```{r one-fit}
train <- example_system$time <= 1
fit <- FitAdaStableNet(
  Y = example_system$Y[train, ],
  tt = example_system$time[train],
  nbasis = 14,
  lambda_range = c(-14, 1),
  twoSE = FALSE,
  eigen_real_wald = TRUE,
  eigen_bound = TRUE,
  num_iter = 120,
  n_starts = 1,
  wald_nsteps = 6,
  seed = 101,
  verbose = FALSE
)

summary(fit)
```

## Optimization backends

`backend = "base"` is the portable default and uses finite-difference BFGS
through `optim()`. The optional `backend = "torch"` evaluates the same profiled
modal loss with float64 tensors and automatic differentiation, optimizes with
Adam, and can refine the best iterate with L-BFGS. Install it once with
`install.packages("torch")` followed by `torch::install_torch()`.

```{r torch-fit, eval=FALSE}
fit_torch <- FitAdaStableNet(
  Y = example_system$Y[train, ],
  tt = example_system$time[train],
  nbasis = 14,
  backend = "torch",
  torch_device = "auto", # CUDA when available, otherwise CPU
  lr = 0.01,
  torch_refine = TRUE,
  torch_refine_iter = 20,
  num_iter = 120,
  seed = 101,
  verbose = FALSE
)

fit_torch$control$backend
fit_torch$AdaEigenStableNet$Eigen_Bound$diagnostics[
  c("device", "dtype", "gradient")
]
```

Use `backend = "auto"` to select Torch only when its package and runtime are
available. The resolved backend, device, data type, convergence history, and
gradient method are retained in the fit diagnostics for reproducibility.

```{r one-forecast, fig.cap="Noise-free trajectory and stable AdaStableNet forecast."}
forecast <- predict(fit, example_system$time, branch = "stable")
matplot(example_system$time, example_system$X, type = "l", lty = 1,
        col = "grey65", xlab = "Time", ylab = "State")
matlines(example_system$time, forecast, lty = 2, col = "#0072B2")
abline(v = 1, lty = 3)
```

## Independent Monte Carlo experiment

The compact experiment runs during package checks. Set the environment variable
`ADASTABLENET_FULL_VIGNETTE=true` before rendering to increase the replication
count. A paper-scale run should use a job array and preserve one result file per
design condition.

```{r simulation}
n_rep <- if (identical(Sys.getenv("ADASTABLENET_FULL_VIGNETTE"), "true")) 100L else 2L
design <- expand.grid(
  spectrum = c("stable", "marginal", "mixed"),
  replicate = seq_len(n_rep),
  stringsAsFactors = FALSE
)

results <- vector("list", nrow(design) * 4L)
row_id <- 0L

for (i in seq_len(nrow(design))) {
  d <- design[i, ]
  sim <- simulate_adastablenet(
    p = 4,
    n_time = 51,
    time_range = c(0, 2),
    spectrum = d$spectrum,
    sigma = 0.08,
    condition_number = 8,
    seed = 1000 + i
  )
  train <- sim$time <= 1
  fitted_model <- FitAdaStableNet(
    sim$Y[train, ], sim$time[train],
    nbasis = 12, lambda_range = c(-14, 1), twoSE = FALSE,
    num_iter = 80, wald_nsteps = 4,
    seed = 2000 + i, verbose = FALSE
  )

  estimates <- list(
    two_stage = fitted_model$Ode2Stage$Ahat,
    unbounded = coef(fitted_model, "unbounded"),
    wald = coef(fitted_model, "wald"),
    stable = coef(fitted_model, "stable")
  )

  for (method in names(estimates)) {
    A_hat <- estimates[[method]]
    if (method == "two_stage") {
      predicted <- t(vapply(
        sim$time,
        function(ti) as.numeric(expm::expm(ti * A_hat) %*% sim$Y[1, ]),
        numeric(nrow(A_hat))
      ))
    } else {
      predicted <- predict(fitted_model, sim$time, branch = method)
    }
    future <- sim$time > 1
    row_id <- row_id + 1L
    results[[row_id]] <- data.frame(
      spectrum = d$spectrum,
      replicate = d$replicate,
      method = method,
      A_relative_error = sqrt(sum((A_hat - sim$A)^2) / sum(sim$A^2)),
      spectral_abscissa_error = abs(
        max(Re(eigen(A_hat)$values)) - max(Re(eigen(sim$A)$values))
      ),
      stable_estimate = max(Re(eigen(A_hat)$values)) <= 1e-7,
      future_RMSE = sqrt(mean((predicted[future, ] - sim$X[future, ])^2))
    )
  }
}

simulation_results <- do.call(rbind, results)
aggregate(
  cbind(A_relative_error, spectral_abscissa_error, future_RMSE) ~
    spectrum + method,
  data = simulation_results,
  FUN = median
)
```

```{r simulation-plot, fig.cap="Future forecast error across independent estimator replications."}
boxplot(
  log10(future_RMSE) ~ interaction(spectrum, method, sep = "\n"),
  data = simulation_results,
  las = 2,
  ylab = expression(log[10]("future RMSE")),
  xlab = "Spectrum and method",
  col = "grey90"
)
```

## Paper-scale protocol

The complete sourceable runner is installed with the package. A two-replication
check can be launched with:

```{r source-runner, eval=FALSE}
Sys.setenv(
  ADASTABLENET_N_REP = 2,
  ADASTABLENET_NOISE_SD = "0.30",
  ADASTABLENET_BACKEND = "base",
  ADASTABLENET_OUTPUT_DIR = file.path(
    getwd(), "adastablenet-simulation-results"
  )
)
source(system.file(
  "scripts", "run-simulation-study.R",
  package = "AdaStableNet"
))
```

It writes one checkpoint after every fitted replication and can therefore be
resumed safely by sourcing the script again with the same output directory.

The journal simulation should extend the executable experiment along the
following prespecified axes:

| Factor | Recommended levels |
|---|---|
| Spectrum | stable, marginal, mixed, near-boundary, unstable |
| Dimension (p) | 15 and 16 for the primary sparse designs |
| Observation times | 31, 61, 101 |
| Noise SD | 0.05, 0.15, 0.30 |
| Matrix structure | sparse (primary), dense (sensitivity analysis) |
| Optimization backend | base, torch; report device and package version |
| Sparse loading threshold | 0.25, 0.50, 0.65 |
| Dense loading condition number | 1, 10, 50 |
| Replications per retained condition | at least 500; 1000 for the main conditions |

The primary comparisons are the two-stage initializer and the unbounded,
Wald-screened, and stable AdaStableNet branches. External methods such as STSLS
should be included only through a portable, versioned implementation with the
same input data and tuning protocol.

Each result record should contain matrix error, matched eigenvalue error,
spectral-abscissa error, stability classification, in-range trajectory error,
future prediction error, convergence code, loading rank and condition number,
proportions of structural zeros in (Q) and (A), optimization backend and device,
and elapsed time.
Oracle-initial-state forecasts and estimated-initial-state forecasts answer
different questions and must be reported separately.

For Wald inference, additional null and alternative conditions should report
type-I error, power, interval coverage, false discovery rate, and edge-selection
accuracy. Monte Carlo standard errors and paired method differences should
accompany all summary tables.
