---
title: "Simulation Study for AdaStableNet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study for AdaStableNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
set.seed(2026)
```

## Purpose

This vignette provides an executable version of the simulation framework used
to assess AdaStableNet. The main estimand is the matrix (A) in

\[
X'(t)=A X(t), \qquad X(t)=\exp\{A(t-t_0)\}X(t_0).
\]

The experiment separates three questions:

1. How accurately is (A) recovered?
2. Does the estimated system have the intended stability behavior?
3. How well does the estimate forecast beyond the observation interval?

Unlike an experiment that fits one trajectory and evaluates many initial
conditions, each Monte Carlo replication below generates new data and refits
every estimator. This measures estimator bias and variability rather than only
transfer across initial conditions.

```{r libraries, message=FALSE}
library(AdaStableNet)
```

## Data-generating systems

`simulate_adastablenet()` constructs a real modal matrix (J), a loading matrix
(Q) with controlled condition number, and (A=QJQ^{-1}). The available
spectral regimes are:

- `stable`: every real part is negative;
- `marginal`: all real parts are zero;
- `mixed`: marginal oscillations plus stable modes;
- `near_boundary`: negative real parts close to zero;
- `unstable`: positive real parts, used as a misspecification check.

The two experiments in the original ODE eigen-bound notebook correspond most
closely to a marginal even-dimensional system and a mixed odd-dimensional
system. They remain important boundary cases because a strict stability
constraint can bias a truly marginal process unless the Wald screen retains a
zero real part.

```{r one-system}
example_system <- simulate_adastablenet(
  p = 4,
  n_time = 61,
  time_range = c(0, 2),
  spectrum = "mixed",
  sigma = 0.08,
  condition_number = 8,
  seed = 11
)

sort(eigen(example_system$A)$values)
```

Only observations on ([0,1]) are used for estimation; ((1,2]) is reserved for
future prediction.

```{r one-fit}
train <- example_system$time <= 1
fit <- FitAdaStableNet(
  Y = example_system$Y[train, ],
  tt = example_system$time[train],
  nbasis = 14,
  lambda_range = c(-14, 1),
  twoSE = FALSE,
  eigen_real_wald = TRUE,
  eigen_bound = TRUE,
  num_iter = 120,
  n_starts = 1,
  wald_nsteps = 6,
  seed = 101,
  verbose = FALSE
)

summary(fit)
```

```{r one-forecast, fig.cap="Noise-free trajectory and stable AdaStableNet forecast."}
forecast <- predict(fit, example_system$time, branch = "stable")
matplot(example_system$time, example_system$X, type = "l", lty = 1,
        col = "grey65", xlab = "Time", ylab = "State")
matlines(example_system$time, forecast, lty = 2, col = "#0072B2")
abline(v = 1, lty = 3)
```

## Independent Monte Carlo experiment

The compact experiment runs during package checks. Set the environment variable
`ADASTABLENET_FULL_VIGNETTE=true` before rendering to increase the replication
count. A paper-scale run should use a job array and preserve one result file per
design cell.

```{r simulation}
n_rep <- if (identical(Sys.getenv("ADASTABLENET_FULL_VIGNETTE"), "true")) 100L else 2L
design <- expand.grid(
  spectrum = c("stable", "marginal", "mixed"),
  replicate = seq_len(n_rep),
  stringsAsFactors = FALSE
)

results <- vector("list", nrow(design) * 4L)
row_id <- 0L

for (i in seq_len(nrow(design))) {
  d <- design[i, ]
  sim <- simulate_adastablenet(
    p = 4,
    n_time = 51,
    time_range = c(0, 2),
    spectrum = d$spectrum,
    sigma = 0.08,
    condition_number = 8,
    seed = 1000 + i
  )
  train <- sim$time <= 1
  fitted_model <- FitAdaStableNet(
    sim$Y[train, ], sim$time[train],
    nbasis = 12, lambda_range = c(-14, 1), twoSE = FALSE,
    num_iter = 80, wald_nsteps = 4,
    seed = 2000 + i, verbose = FALSE
  )

  estimates <- list(
    two_stage = fitted_model$Ode2Stage$Ahat,
    unbounded = coef(fitted_model, "unbounded"),
    wald = coef(fitted_model, "wald"),
    stable = coef(fitted_model, "stable")
  )

  for (method in names(estimates)) {
    A_hat <- estimates[[method]]
    if (method == "two_stage") {
      predicted <- t(vapply(
        sim$time,
        function(ti) as.numeric(expm::expm(ti * A_hat) %*% sim$Y[1, ]),
        numeric(nrow(A_hat))
      ))
    } else {
      predicted <- predict(fitted_model, sim$time, branch = method)
    }
    future <- sim$time > 1
    row_id <- row_id + 1L
    results[[row_id]] <- data.frame(
      spectrum = d$spectrum,
      replicate = d$replicate,
      method = method,
      A_relative_error = sqrt(sum((A_hat - sim$A)^2) / sum(sim$A^2)),
      spectral_abscissa_error = abs(
        max(Re(eigen(A_hat)$values)) - max(Re(eigen(sim$A)$values))
      ),
      stable_estimate = max(Re(eigen(A_hat)$values)) <= 1e-7,
      future_RMSE = sqrt(mean((predicted[future, ] - sim$X[future, ])^2))
    )
  }
}

simulation_results <- do.call(rbind, results)
aggregate(
  cbind(A_relative_error, spectral_abscissa_error, future_RMSE) ~
    spectrum + method,
  data = simulation_results,
  FUN = median
)
```

```{r simulation-plot, fig.cap="Future forecast error across independent estimator replications."}
boxplot(
  log10(future_RMSE) ~ interaction(spectrum, method, sep = "\n"),
  data = simulation_results,
  las = 2,
  ylab = expression(log[10]("future RMSE")),
  xlab = "Spectrum and method",
  col = "grey90"
)
```

## Paper-scale protocol

The journal simulation should extend the executable experiment along the
following prespecified axes:

| Factor | Recommended levels |
|---|---|
| Spectrum | stable, marginal, mixed, near-boundary, unstable |
| Dimension (p) | 4, 8, 16 |
| Observation times | 31, 61, 101 |
| Noise SD | 0.05, 0.15, 0.30 |
| Loading condition number | 1, 10, 50 |
| Replications per retained cell | at least 500 |

The primary comparisons are the two-stage initializer and the unbounded,
Wald-screened, and stable AdaStableNet branches. External methods such as STSLS
should be included only through a portable, versioned implementation with the
same input data and tuning protocol.

Each result record should contain matrix error, matched eigenvalue error,
spectral-abscissa error, stability classification, in-range trajectory error,
future prediction error, convergence code, loading rank and condition number,
and elapsed time. Oracle-initial-state forecasts and estimated-initial-state
forecasts answer different questions and must be reported separately.

For Wald inference, additional null and alternative cells should report
type-I error, power, interval coverage, false discovery rate, and edge-selection
accuracy. Monte Carlo standard errors and paired method differences should
accompany all summary tables.
