---
title: "Simulation and Network-Recovery Example for AdaStableNet"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation and Network-Recovery Example for AdaStableNet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 7,
                      fig.height = 4.5)
set.seed(2026)
```

## Purpose

This vignette provides a compact executable example for checking AdaStableNet
on simulated data. The main estimand is the matrix (A) in

\[
X'(t)=A X(t), \qquad X(t)=\exp\{A(t-t_0)\}X(t_0).
\]

This is a standard time-domain ODE setting. Observations are multivariate state
measurements collected at ordered times; they are not cells, and the time
variable is observed directly rather than inferred as pseudotime.

The experiment separates three questions:

1. How accurately is (A) recovered?
2. Does the estimated system have the intended stability behavior?
3. How well does the estimate forecast beyond the observation interval?

Unlike an experiment that fits one trajectory and evaluates many initial
conditions, each Monte Carlo replication below generates new data and refits
every estimator. This measures estimator bias and variability rather than only
transfer across initial conditions.

```{r libraries, message=FALSE}
library(AdaStableNet)
```

## Data-generating systems

`simulate_adastablenet()` constructs a real modal matrix (J), a loading matrix
(Q), and (A=QJQ^{-1}). By default, `matrix_structure = "sparse"` reproduces
the eigen-bound notebook construction: two small orthogonal blocks are embedded
in an identity matrix, multiplied by a randomly permuted local rotation, and
entries below the median nonzero magnitude are set to zero. This produces a
sparse, non-orthogonal (Q) and a sparse (A). Set
`matrix_structure = "dense"` to use a dense loading matrix with an exact target
condition number.

The realized sparsity is recorded in `Q_sparsity` and `A_sparsity`. In very
small systems, the embedded blocks can cover most coordinates, so the realized
proportion of zeros in (A) may be lower than in the original (p=15) and (p=16)
designs.

The available spectral regimes are:

- `stable`: every real part is negative;
- `marginal`: all real parts are zero;
- `mixed`: marginal oscillations plus a stable real mode for odd dimensions;
  even dimensions alternate marginal and stable oscillatory modes;
- `near_boundary`: negative real parts close to zero;
- `unstable`: positive real parts, used as a misspecification check.

The two experiments in the original ODE eigen-bound notebook correspond most
closely to a marginal even-dimensional system and a mixed odd-dimensional
system. They remain important boundary cases because a strict stability
constraint can bias a truly marginal process unless the Wald screen retains a
zero real part.

```{r one-system}
sparse_system <- simulate_adastablenet(
  p = 16,
  n_time = 61,
  time_range = c(0, 2),
  spectrum = "marginal",
  sigma = 0.08,
  condition_number = 5,
  seed = 777
)

sort(eigen(sparse_system$A)$values)
sparse_system[c("matrix_structure", "Q_sparsity", "A_sparsity",
                "condition_number")]

example_system <- simulate_adastablenet(
  p = 4,
  n_time = 61,
  time_range = c(0, 2),
  spectrum = "mixed",
  sigma = 0.08,
  condition_number = 8,
  seed = 11
)
```

```{r sparsity-pattern, fig.cap="Nonzero pattern of the sparse system matrix A."}
image(
  x = seq_len(nrow(sparse_system$A)),
  y = seq_len(ncol(sparse_system$A)),
  z = abs(sparse_system$A) >= 1e-12,
  col = c("white", "#0072B2"),
  xlab = "Row", ylab = "Column", axes = FALSE
)
axis(1)
axis(2)
box()
```

Only observations on ([0,1]) are used for estimation; ((1,2]) is reserved for
future prediction.

```{r one-fit}
train <- example_system$time <= 1
fit <- FitAdaStableNet(
  Y = example_system$Y[train, ],
  tt = example_system$time[train],
  nbasis = 14,
  lambda_range = c(-14, 1),
  twoSE = FALSE,
  eigen_real_wald = TRUE,
  eigen_bound = TRUE,
  num_iter = 120,
  n_starts = 1,
  wald_nsteps = 6,
  seed = 101,
  verbose = FALSE
)

summary(fit)
```

## Coefficient-Wald sparsification and network recovery

The modal `wald` branch and coefficient sparsification are different steps.
The former screens modal real parts before constrained fitting. The latter
tests entries of the assembled coefficient matrix. For the network, only
off-diagonal entries are eligible edges: under `X'(t)=AX(t)`, `A[i,j]` is the
directed effect `j -> i`. Diagonal coefficients are self-dynamics rather than
edges.

```{r one-network}
coefficient_network <- AdaStableNet_WaldNetwork(
  fit,
  example_system$Y[train, ],
  example_system$time[train],
  branch = "stable",
  method = "BH",
  alpha = 0.05,
  diagonal = "keep",
  nsteps = 6,
  stability_horizon = 2,
  stability_n_grid = 21
)

coefficient_network$adjacency
coefficient_network$A_network
coefficient_network$A_sparse
head(coefficient_network$edge_table)
```

`adjacency` is binary and `A_network` is the signed, weighted graph; both have
zero diagonals. `A_sparse` is the matrix that can be inserted back into the ODE
and therefore retains diagonal self-dynamics by default. Thresholding is not a
spectral operation and can change stability, so the object records diagnostics
for both the fitted and sparsified matrices. The primary fitted trajectory
should continue to use the unthresholded branch; prediction under `A_sparse`
is reported as a sensitivity analysis.

In simulation, edge support is known. ROC AUC and precision-recall AUC must be
computed by sweeping the continuous absolute Wald score over all
off-diagonal coefficients, not from the single BH network. The prespecified
BH network at `alpha = 0.05` supplies power, false-positive rate, false-
discovery proportion, precision, F1, and sign recovery. For observed data,
where support is unknown, ROC claims are replaced by BH-selected edges,
bootstrap or replicate selection frequency, and sign reproducibility.

## Optimization backends

`backend = "base"` is the portable default and uses finite-difference BFGS
through `optim()`. The optional `backend = "torch"` evaluates the same profiled
modal loss with float64 tensors and automatic differentiation, optimizes with
Adam, and can refine the best iterate with L-BFGS. Install it once with
`install.packages("torch")` followed by `torch::install_torch()`.

```{r torch-fit, eval=FALSE}
fit_torch <- FitAdaStableNet(
  Y = example_system$Y[train, ],
  tt = example_system$time[train],
  nbasis = 14,
  backend = "torch",
  torch_device = "auto", # CUDA when available, otherwise CPU
  lr = 0.01,
  torch_refine = TRUE,
  torch_refine_iter = 20,
  num_iter = 120,
  seed = 101,
  verbose = FALSE
)

fit_torch$control$backend
fit_torch$AdaEigenStableNet$Eigen_Bound$diagnostics[
  c("device", "dtype", "gradient")
]
```

Use `backend = "auto"` to select Torch only when its package and runtime are
available. The resolved backend, device, data type, convergence history, and
gradient method are retained in the fit diagnostics for reproducibility.

```{r one-forecast, fig.cap="Noise-free trajectory and stable AdaStableNet forecast."}
forecast <- predict(fit, example_system$time, branch = "stable")
matplot(example_system$time, example_system$X, type = "l", lty = 1,
        col = "grey65", xlab = "Time", ylab = "State")
matlines(example_system$time, forecast, lty = 2, col = "#0072B2")
abline(v = 1, lty = 3)
```

## Compact Monte Carlo example

The deliberately small experiment below illustrates independent refitting while
remaining appropriate for package documentation. Large simulation designs and
their checkpoints belong in a separate analysis repository.

```{r simulation}
n_rep <- 2L
design <- expand.grid(
  spectrum = c("stable", "marginal", "mixed"),
  replicate = seq_len(n_rep),
  stringsAsFactors = FALSE
)

results <- vector("list", nrow(design) * 4L)
row_id <- 0L

for (i in seq_len(nrow(design))) {
  d <- design[i, ]
  sim <- simulate_adastablenet(
    p = 4,
    n_time = 51,
    time_range = c(0, 2),
    spectrum = d$spectrum,
    sigma = 0.08,
    condition_number = 8,
    seed = 1000 + i
  )
  train <- sim$time <= 1
  fitted_model <- FitAdaStableNet(
    sim$Y[train, ], sim$time[train],
    nbasis = 12, lambda_range = c(-14, 1), twoSE = FALSE,
    fit_ode2fd = TRUE,
    num_iter = 80, wald_nsteps = 4,
    seed = 2000 + i, verbose = FALSE
  )

  estimates <- list(
    two_stage = fitted_model$Ode2Stage$Ahat,
    unbounded = coef(fitted_model, "unbounded"),
    wald = coef(fitted_model, "wald"),
    stable = coef(fitted_model, "stable")
  )

  for (method in names(estimates)) {
    A_hat <- estimates[[method]]
    predicted <- predict(fitted_model, sim$time, branch = method)
    future <- sim$time > 1
    row_id <- row_id + 1L
    results[[row_id]] <- data.frame(
      spectrum = d$spectrum,
      replicate = d$replicate,
      method = method,
      A_relative_error = sqrt(sum((A_hat - sim$A)^2) / sum(sim$A^2)),
      spectral_abscissa_error = abs(
        max(Re(eigen(A_hat)$values)) - max(Re(eigen(sim$A)$values))
      ),
      stable_estimate = max(Re(eigen(A_hat)$values)) <= 1e-7,
      future_RMSE = sqrt(mean((predicted[future, ] - sim$X[future, ])^2))
    )
  }
}

simulation_results <- do.call(rbind, results)
aggregate(
  cbind(A_relative_error, spectral_abscissa_error, future_RMSE) ~
    spectrum + method,
  data = simulation_results,
  FUN = median
)
```

```{r simulation-plot, fig.cap="Future forecast error across independent estimator replications."}
boxplot(
  log10(future_RMSE) ~ interaction(spectrum, method, sep = "\n"),
  data = simulation_results,
  las = 2,
  ylab = expression(log[10]("future RMSE")),
  xlab = "Spectrum and method",
  col = "grey90"
)
```

## Stability, coefficient inference, and identifiability

The `stable` branch is the constrained optimization, not the coefficient Wald
test. Its default constraint controls the spectral abscissa. Because spectral
stability does not exclude transient growth for nonnormal matrices, use
`stability_diagnostics()` to record the numerical abscissa and the maximum
finite-horizon matrix-exponential norm.

`AdaStableNet_WaldTest()` implements the coefficientwise tensor/integration
calculation based on the total trajectory Fisher information of Wu et al.
(2019). Apply it separately to the unbounded and constrained branches and
evaluate false-positive rate, false-discovery proportion, power, and sign
recovery. The usual normal reference is a working approximation for a
boundary-constrained estimate.

When `ode.ident` is installed, `identifiability_diagnostics(A, x0)` exposes the
Qiu et al. initial-condition identifiability score (ICIS). When ICIS is used in
a research analysis, retain the full population of generated initial conditions;
retaining only the largest ICIS values would optimistically select the easiest
trajectories.
